k-mer-based GWAS enhances the discovery of causal variants and candidate genes in soybean

### 2023-03-31 additional_files/manuscript.tex
Fix citations by adding author names where required

Alternative approaches that allow one to assess a large spectrum
of variants in GWAS without genotyping SVs at population scale could be useful.
\cite{voichek2020} developed a \emph{k}-mer-based GWAS approach that may
represent such an alternative method. This approach relies on the presence or
absence of \emph{k}-mers observed in sequence reads in place of variant genotypes for use in
GWAS. Once a presence/absence table of \emph{k}-mers has been generated for the
population under study, this table can be used in association analyses
similarly to SNP or SV genotype calls. One advantage of using \emph{k}-mers is
that they can act as a molecular signature for any type of variant as long as
these variants result in presence/absence of \emph{k}-mers of a given length.
Another advantage is that \emph{k}-mers are not tied to a specific genomic
location and can thus be used to query associations between genotype and
phenotype at genomic locations that are not included in a reference genome.  A
few studies have applied this \emph{k}-mer-based GWAS approach
\cite[e.g.][]{tripodi2021, colque2021} and other similar approaches have been
developed \cite{rahman2018, he2021}, but similarly to the use of SVs in GWAS,
such approaches have not been widely adopted yet.

@@

\cite{zabala2007} identified the flavonoid 3’5’-hydroxylase (F3’5’H) gene
Glyma.13g\-072100 as associated with the \textit{W1} locus for flower color. In
Williams82 and other accessions with white flowers, a 65-bp insertion into the
third exon of the gene results in a premature stop codon that renders the
F3’5’H enzyme non-functional and prevents pigmentation of the flower. All
GWAS approaches detected a signal overlapping the gene at this locus (Additional file 1: Figures
S1 and S2). The most
significant SV and \textit{k}-mers corresponded to the known causal variant
(Figure \ref{flower-color-main-figure}a-b and Additional file 1: S3c-d).
In particular, the assembly of the reads containing the most significant \textit{k}-mers
and the alignment of the resulting haplotypes showed that these \textit{k}-mers
directly tagged the causal variant (Figure \ref{flower-color-main-figure}c-e).
As the catalogue of variants analyzed by Platypus contains only SNPs and small
indels, the causal variant could not be captured by this approach. Still, SNPs
and indels located within the gene sequence were among the most significantly
associated ones (Additional file 1: Figure S3b).

@@

\cite{zabala2003} identified Glyma.06g202300 as the gene associated with the
\textit{T} locus for pubescence color and found a 1-bp deletion as a putative
causal variant at this locus. For all three GWAS approaches, we found a signal
overlapping this gene (Additional file 1: Figure S5). In
particular, the regions defined by the 1\% most associated variants
identified by the SNP- and \textit{k}-mer-based approaches included
Glyma.06g202300 (Additional file 1: Figure S6). While the most
significant \textit{k}-mers did occur within the sequence of this gene, they
did not correspond to the documented causal variant (Additional file 1: Figure
S7d).  Still, the fourth and fifth most strongly
associated \textit{k}-mers corresponded to this causal variant (Additional file 1: Figures
S7d and S8). We
observed that not all accessions with gray pubescence harbored the deleted
nucleotide (Additional file 1: Figure S8c), which could indicate
either that other causal variants are resulting in a non-functional F3’H in
this population, or that the causal variant lies elsewhere. As for SNP-based GWAS,
the most significant variant within the sequence of Glyma.06g202300 was the
causal variant, however it was not the most significantly associated variant
overall (Additional file 1: Figures S6b and
S7b). SV-based GWAS did not detect the causal variant
at this locus, as expected from its size.

@@

We conducted a second GWAS analysis contrasting accessions with yellow and
green seed coats in order to target the \textit{G} locus. Accordingly, this
analysis found strong signals overlapping the \emph{G} locus using all three
approaches (Additional file 1: Figure S14).
\cite{wang2018} identified Glyma.01g198500 as the gene associated with
this locus and found an A>G SNP affecting transcript splicing as the causal
variant. Consistently with their results, we identified the two
most significant \emph{k}-mers and the second most significant marker from the
SNP-based analysis as corresponding to that causal variant (Figure
\ref{seed-coat-color-main-figure} and Additional file 1: Figures
S15 and
S16). Unsurprisingly, SV-based GWAS
did not identify the causal variant given its type.

@@

While all GWAS approaches found signals near the Glyma.19g194300 gene
associated with the \textit{Dt1} locus for stem termination type, the SNP- and
\textit{k}-mer-based analyses performed best since the region defined by their
most highly associated markers (top 1\%) included the causal gene (Additional file 1: Figures
S17 and S18).
We did identify one of the SNPs suggested as putative causal variants for this
locus by \cite{tian2010} and \cite{liu2010}, however this SNP was only the
135\textsuperscript{th} most associated marker found by Platypus and the
12\textsuperscript{th} most associated \emph{k}-mer (Additional file 1: Figures
S19 and S20).
Our results are consistent with previous findings in suggesting that there is
more than one causal variant at the \textit{Dt1} locus \citep{liu2010,
tian2010}. Indeed, a few cultivars bearing the allele associated with
\emph{dt1} showed indeterminate or semi-determinate phenotypes and some
cultivars showed a determinate phenotype without bearing the allele associated
with \emph{dt1} (Additional file 1: Figure S20c). The existence of
several causal variants might explain why our analyses did not identify a
causal variant more clearly.

@@

We conducted a GWAS analysis that only compared accessions with black and brown
hila in an attempt to detect stronger signals at the \textit{R} locus.  We
accordingly detected a signal overlapping this locus using all approaches, but
only SNP- and \textit{k}-mer-based GWAS detected signals overlapping the
Glyma.09g235100 gene associated with \textit{R} (Additional file 1: Figures
S21 and S22).
\cite{gillman2011} identified this R2R3 MYB transcription factor as the
molecular basis for the \textit{R} locus and documented four different
loss-of-function alleles linked to brown hilum color. Although our analyses did
detect strong signals in the vicinity of this gene, the most associated markers
were not located in the body of the gene (Additional file 1: Figure
S22).  Interestingly, the most significantly
associated \emph{k}-mers detected by our analysis mapped 2 kb upstream of
another gene (Glyma.09g234900) annotated as a MYB transcription factor and
putatively involved in the regulation of anthocyanin biosynthesis. This gene as
well as Glyma.09g235100 and two others are part of a tandem array of four R2R3
MYB genes considered as candidates by \cite{gillman2011}, and of which
Glyma.09g235100 was the only one expressed in the seed coat. Given the
convincing evidence provided by \cite{gillman2011}, it is most likely that
Glyma.09g235100 is indeed the gene associated with the \emph{R} locus and that
our failure to identify the most significantly associated \emph{k}-mers as
causal variants is due to the coexistence of several variants that can cause
loss of function at the \emph{R} locus.  Still, we observed two causal variants
previously documented by \cite{gillman2011} in our dataset (Additional file 1: Figures
S23 and S24).

@@

We conducted a GWAS analysis comparing accessions with normal and semi-sparse
pubes\-cence in order to detect loci associated with pubescence density. This analysis
detected signals at the \textit{Ps} locus using all three approaches (Figure
\ref{pubescence-density-main-figure}a-b and Additional file 1: Figures S25, S26
and S27). The causal
variant at this recently cloned locus is a copy number variant (CNV) overlapping the
Glyma.12g187200 gene, with higher copy number resulting in reduced pubescence \citep{liu2020ps}.
We identified a region of $\sim$42 kb that varied in copy number across
the population and was associated with the trait (Figure
\ref{pubescence-density-main-figure}c), and therefore possibly corresponded to the
25.6-kb region identified as the causal CNV by \cite{liu2020ps}. We found that
reads containing the most significantly associated \textit{k}-mer mapped to either of the two ends
of the CNV (Figure \ref{pubescence-density-main-figure}d). This suggests that
the most associated \textit{k}-mer is indeed linked to the CNV
that is causal at this locus, while \textit{k}-mers within the
CNV region may represent differences in sequence between the copies.

The difference in size between the CNV reported by \cite{liu2020ps} and our
results may be due to the presence of some unresolved sequence (a stretch of
4,635 N nucleotides) over that interval in the reference sequence that we used.
Because our SV filtering process removed any alternate allele containing N nucleotides,
we reasoned that the CNV may have been found by our pipeline but
filtered out later on. Indeed, we found a 42.8-kb duplication corresponding
to the one described above among the genotype calls made by smoove. An association
analysis of that single variant using GAPIT found a \textit{p}-value of 1.9e-23,
which would have made it the most significant SV. In this particular case, the
failure of the SV-based GWAS analysis to find identify the causal variant was due
to our filtering parameters and not to limitations of the SV genotype calls per se.

@@

We conducted a GWAS analysis comparing dull and shiny seed coat phenotypes for
the seed coat luster trait. Seed coat luster has been shown by
\cite{gijzen1999} to be largely caused by the deposition of a hydrophobic
protein (HPS) at the seed surface. Copy number variation of this gene at the
\textit{B} locus was later shown to explain variation in luster, although
sequence variation in the HPS sequence may also play a role \citep{gijzen2006}.
Using a sequencing depth-based analysis, we identified a 31-kb region spanning
positions 9,386,109-9,417,431 on chromosome Gm15 that exhibited variation in
copy number across accessions. This CNV overlapped four genes annotated as
containing HPS domains and therefore likely represented the causal variant
(Additional file 1: Figures S28 and
S29).  We found signals overlapping this
CNV using all approaches, and the most significantly associated \textit{k}-mer
was notably located within the CNV region (Additional file 1: Figures
S28 and
S30). However, we were not able to
directly link any of the variants or \textit{k}-mers identified to the causal
CNV.

@@

Both the \textit{L1} and the \textit{L2} loci for pod color have yet to be
cloned. While \cite{he2015} proposed the gene Glyma.19g101700 as likely
corresponding to \textit{L1} based on fine mapping and expression analysis, our
results as well as those reported by \cite{bandillo2017} do not support this
gene as a candidate due to its location $\sim$3 Mb away from GWAS signals. Our
analyses based on \textit{k}-mers and SNPs found signals overlapping 14 and
6 genes, respectively (Additional file 1: S31 and
S32). Of these, a MATE transporter (Glyma.19g120300)
appears as a prime candidate as the most significant \emph{k}-mer was observed
within its sequence and it is highly expressed in pods according to data hosted
on SoyBase \citep{grant2010, severin2010}. Moreover, the role of some MATE
proteins in transporting flavonoids has been demonstrated \citep{chen2015}.

@@

The two known loci for pubescence form in soybean, \textit{Pa1} and
\textit{Pa2}, have also not been cloned yet. Nevertheless, \cite{gilbert2017}
has suggested Glyma.12g213900 as a candidate for \textit{Pa1} based on data
published by \cite{bandillo2017}. This gene encodes a MYB transcription factor
annotated as playing a role in trichome branching. Indeed, the most significant
signals found by all three approaches were located near this gene (Figure
\ref{pubescence-form-main-figure}a-b, and Additional file 1: Figures S34 and
S35).  The analyses based on SNPs and
\emph{k}-mers identified two highly significant non-synonymous SNPs in
Glyma.12g213900 that were associated with appressed pubescence (Figure
\ref{pubescence-form-main-figure}c-d and Additional file 1: Figures S36 and
S37). The alternative alleles for these two
SNPs co-occurred together in our dataset and may represent causal variants at
the \emph{Pa1} locus. The signals observed at the \emph{Pa2} locus were much
weaker and the signal observed from the \emph{k}-mer GWAS overlapped a total of
20 genes (Additional file 1: Figures S34 and
S38), none of which appeared obviously linked
to pubescence form.

@@

We used corrected dry weight (CDW) as described by \cite{deronne2022} as a
measure of horizontal resistance to \emph{P. sojae}. We found a strong signal
overlapping the genomic region reported by these authors on Gm15 using all three GWAS
approaches (Additional file 1: Figure S40). \cite{deronne2022}
proposed Glyma.15g217100 as a candidate gene for explaining phenotypic
variation at that locus based on a combination of functional annotation and
contrasting gene expression in resistant and susceptible lines four days after
inoculation. Although the signals found by our analyses did overlap
Glyma.15g217100, we found that this gene was located at the distal end of the signal
in a region with markers or \emph{k}-mers showing a weaker
association (Additional file 1: Figure S41). The
large size of the observed signal means that the gene associated with the phenotype could
be located almost anywhere in this region. However, we identified a particularly
interesting region defined by the top 1\% associations in the SNP-based
analysis. This region overlaps only three annotated genes and also contains the
most significant \emph{k}-mer (Additional file 1: Figure
S41). One of the genes in this region is
a calcium ion binding protein with an EF-hand domain (Glyma.15g217700).
Given the important role played by calcium signaling in defense
responses \citep{zhang2014}, we suggest Glyma.15g217700 as an interesting
candidate gene for explaining resistance to \emph{P. sojae} in this population.

@@

As would be expected of any GWAS analysis, the success of the \emph{k}-mer approach
depends on the quality of the phenotypic data and the complexity of the genetic
architecture of the trait. It is clear that the loci for which the
\emph{k}-mer approach was most successful (\emph{W1}, \emph{Td}, \emph{G} and
\emph{Ps}) have a simple genetic architecture and phenotypes that can be easily
determined. For other traits, challenges in determining accurate phenotypes may
result in a decreased ability to identify causal variants or genes. For
example, \cite{bandillo2017} noted that observed phenotypes for stem
termination type in the GRIN database may not correspond to the expected
genotypes at classical loci, and that the intermediate phenotype for seed coat
luster is not always consistent across different environments. For qualitative
traits, the numerical scale used to recode phenotypes for use in GWAS is somewhat
arbitrary and may influence the final results. In the case of quantitative
traits, the usual limitations to the precision and stochastic variation of
phenotypic values will necessarily make the interpretation more difficult than
for simple qualitative traits. Notwithstanding those limitations,
\textit{k}-mer-based GWAS clearly performed better than any of the other
approaches tested here given the same phenotypic data. This is partly due
to the fact that the \textit{k}-mer approach encompasses all variant types.
Moreover, it is likely that the simpler workflow that results from using
\textit{k}-mer presence/absence avoids some of the errors introduced during
variant calling and genotyping, and therefore yields more accurate results.

@@

The complexity of analyzing the results of \emph{k}-mer-based GWAS is another
limitation of this approach. Although the method developed by
\cite{voichek2020} is in itself relatively simple and computationally
inexpensive, downstream analysis and interpretation of the output is
challenging. Indeed, there are no state-of-the-art methods for identifying the
putative genomic coordinates of a \emph{k}-mer, grouping significant
\emph{k}-mers together into coherent signals, and linking them to biologically
meaningful sequence variation. Going forward, mapping the \emph{k}-mers or
reads to a graph-based genome \citep[e.g. using the vg toolkit;][]{siren2021}
may represent an improvement to using a single linear genome.  Grouping
significant \emph{k}-mers by LD instead of by genomic position, as was done for
some of the analyses in this paper, could prove useful in determining how many
loci control a trait. In this study, we used an assembly-based approach similar
to what others have previously done \citep[e.g.][]{voichek2020, rahman2018}.
This approach allowed us to link \emph{k}-mers to their underlying variation
for simple cases involving SNPs and indels, but we were unable to link
significantly associated \emph{k}-mers to more challenging causal variants such
as those explaining the \emph{I} or \emph{B} loci. In this context, developing
approaches to systematically link significant \emph{k}-mers to the sequence
variation that underlies their presence is much needed. Furthermore, finding
ways to limit the number of spurious associations found for some traits, as was
the case for flower color and pubescence color in our study, is also needed in
order to obtain robust results.

@@

Given the current performance of SV genotyping programs, it is not clear
whether using SVs for GWAS provides any advantage over using \emph{k}-mer- or
SNP-based GWAS, or a combination of both.  Indeed, in the single case where the
SV-based analyses identified the causal variant, the \emph{k}-mer-based GWAS
identified it as well. Moreover, SNP-based GWAS provided better guidance
in delimiting the signals found at a given locus because of the much higher
density of markers compared to SVs.  The usefulness of SV-based GWAS compared
to SNP-based GWAS will largely depend on whether SVs can represent variation
that is not otherwise in LD with SNPs. In tomato, \cite{dominguez2020} found
that most transposable element insertions associated with phenotypic variation
in agronomic traits were not tagged by SNPs. On the other hand, in humans,
\cite{maretty2017} found that the majority of SVs were in high LD with SNPs. It
is therefore likely that how well SVs can be tagged by SNPs will depend on the
particular type of SV and on the species. A more efficient approach than
SV-based GWAS may be to conduct \emph{k}-mer-based GWAS and only then link
known SVs to significant \emph{k}-mers.  Alternatively, \textit{k}-mer-based
GWAS can be conducted specifically for \textit{k}-mers that are associated with
SVs \citep{jayakodi2020}.  Previous studies have obtained noteworthy results by
using SVs for GWAS analyses \citep[e.g.][]{zhang2015, akakpo2020,
dominguez2020}, however it would be worthwhile to see whether these results
could be replicated using \emph{k}-mer-based GWAS.

@@

We selected 389 inbred \emph{G. max} accessions based on the availability of
Illumina WGS data and phenotypic data for resistance to the oomycete \emph{P.
sojae} \citep{deronne2022}.  We identified 741 SRA runs corresponding to
Illumina paired-end sequencing data for those 389 accessions \citep[data from][]{zhou2015, valliyodan2016, fang2017, bayer2021}
and extracted paired-end reads using the fastq-dump command v. 2.9.6
(\citeauthor[][\url{https://github.com/ncbi/sra-tools}]{sratoolkit}) with the
option \texttt{-\-split-3}.
Reads that were not paired in the raw sequencing data were not used
for downstream analyses.  The runs downloaded from the SRA and associated
metadata are listed in Additional file 2.

>>>>

Alternative approaches that allow one to assess a large spectrum
of variants in GWAS without genotyping SVs at population scale could be useful.
Voichek and Weigel \cite{voichek2020} developed a \emph{k}-mer-based GWAS approach that may
represent such an alternative method. This approach relies on the presence or
absence of \emph{k}-mers observed in sequence reads in place of variant genotypes for use in
GWAS. Once a presence/absence table of \emph{k}-mers has been generated for the
population under study, this table can be used in association analyses
similarly to SNP or SV genotype calls. One advantage of using \emph{k}-mers is
that they can act as a molecular signature for any type of variant as long as
these variants result in presence/absence of \emph{k}-mers of a given length.
Another advantage is that \emph{k}-mers are not tied to a specific genomic
location and can thus be used to query associations between genotype and
phenotype at genomic locations that are not included in a reference genome.  A
few studies have applied this \emph{k}-mer-based GWAS approach
\cite[e.g.][]{tripodi2021, colque2021} and other similar approaches have been
developed \cite{rahman2018, he2021}, but similarly to the use of SVs in GWAS,
such approaches have not been widely adopted yet.

@@

Zabala and Vodkin \cite{zabala2007} identified the flavonoid 3’5’-hydroxylase (F3’5’H) gene
Glyma.13g\-072100 as associated with the \textit{W1} locus for flower color. In
Williams82 and other accessions with white flowers, a 65-bp insertion into the
third exon of the gene results in a premature stop codon that renders the
F3’5’H enzyme non-functional and prevents pigmentation of the flower. All
GWAS approaches detected a signal overlapping the gene at this locus (Additional file 1: Figures
S1 and S2). The most
significant SV and \textit{k}-mers corresponded to the known causal variant
(Figure \ref{flower-color-main-figure}a-b and Additional file 1: S3c-d).
In particular, the assembly of the reads containing the most significant \textit{k}-mers
and the alignment of the resulting haplotypes showed that these \textit{k}-mers
directly tagged the causal variant (Figure \ref{flower-color-main-figure}c-e).
As the catalogue of variants analyzed by Platypus contains only SNPs and small
indels, the causal variant could not be captured by this approach. Still, SNPs
and indels located within the gene sequence were among the most significantly
associated ones (Additional file 1: Figure S3b).

@@

Zabala and Vodkin \cite{zabala2003} identified Glyma.06g202300 as the gene associated with the
\textit{T} locus for pubescence color and found a 1-bp deletion as a putative
causal variant at this locus. For all three GWAS approaches, we found a signal
overlapping this gene (Additional file 1: Figure S5). In
particular, the regions defined by the 1\% most associated variants
identified by the SNP- and \textit{k}-mer-based approaches included
Glyma.06g202300 (Additional file 1: Figure S6). While the most
significant \textit{k}-mers did occur within the sequence of this gene, they
did not correspond to the documented causal variant (Additional file 1: Figure
S7d).  Still, the fourth and fifth most strongly
associated \textit{k}-mers corresponded to this causal variant (Additional file 1: Figures
S7d and S8). We
observed that not all accessions with gray pubescence harbored the deleted
nucleotide (Additional file 1: Figure S8c), which could indicate
either that other causal variants are resulting in a non-functional F3’H in
this population, or that the causal variant lies elsewhere. As for SNP-based GWAS,
the most significant variant within the sequence of Glyma.06g202300 was the
causal variant, however it was not the most significantly associated variant
overall (Additional file 1: Figures S6b and
S7b). SV-based GWAS did not detect the causal variant
at this locus, as expected from its size.

@@

We conducted a second GWAS analysis contrasting accessions with yellow and
green seed coats in order to target the \textit{G} locus. Accordingly, this
analysis found strong signals overlapping the \emph{G} locus using all three
approaches (Additional file 1: Figure S14).
Wang et al. \cite{wang2018} identified Glyma.01g198500 as the gene associated with
this locus and found an A>G SNP affecting transcript splicing as the causal
variant. Consistently with their results, we identified the two
most significant \emph{k}-mers and the second most significant marker from the
SNP-based analysis as corresponding to that causal variant (Figure
\ref{seed-coat-color-main-figure} and Additional file 1: Figures
S15 and
S16). Unsurprisingly, SV-based GWAS
did not identify the causal variant given its type.

@@

While all GWAS approaches found signals near the Glyma.19g194300 gene
associated with the \textit{Dt1} locus for stem termination type, the SNP- and
\textit{k}-mer-based analyses performed best since the region defined by their
most highly associated markers (top 1\%) included the causal gene (Additional file 1: Figures
S17 and S18).
We did identify one of the SNPs suggested as putative causal variants for this
locus by Tian et al. \cite{tian2010} and Liu et al. \cite{liu2010}, however this SNP was only the
135\textsuperscript{th} most associated marker found by Platypus and the
12\textsuperscript{th} most associated \emph{k}-mer (Additional file 1: Figures
S19 and S20).
Our results are consistent with previous findings in suggesting that there is
more than one causal variant at the \textit{Dt1} locus \citep{liu2010,
tian2010}. Indeed, a few cultivars bearing the allele associated with
\emph{dt1} showed indeterminate or semi-determinate phenotypes and some
cultivars showed a determinate phenotype without bearing the allele associated
with \emph{dt1} (Additional file 1: Figure S20c). The existence of
several causal variants might explain why our analyses did not identify a
causal variant more clearly.

@@

We conducted a GWAS analysis that only compared accessions with black and brown
hila in an attempt to detect stronger signals at the \textit{R} locus.  We
accordingly detected a signal overlapping this locus using all approaches, but
only SNP- and \textit{k}-mer-based GWAS detected signals overlapping the
Glyma.09g235100 gene associated with \textit{R} (Additional file 1: Figures
S21 and S22).
Gillman et al. \cite{gillman2011} identified this R2R3 MYB transcription factor as the
molecular basis for the \textit{R} locus and documented four different
loss-of-function alleles linked to brown hilum color. Although our analyses did
detect strong signals in the vicinity of this gene, the most associated markers
were not located in the body of the gene (Additional file 1: Figure
S22).  Interestingly, the most significantly
associated \emph{k}-mers detected by our analysis mapped 2 kb upstream of
another gene (Glyma.09g234900) annotated as a MYB transcription factor and
putatively involved in the regulation of anthocyanin biosynthesis. This gene as
well as Glyma.09g235100 and two others are part of a tandem array of four R2R3
MYB genes considered as candidates by Gillman et al. \cite{gillman2011}, and of which
Glyma.09g235100 was the only one expressed in the seed coat. Given the
convincing evidence provided by Gillman et al. \cite{gillman2011}, it is most likely that
Glyma.09g235100 is indeed the gene associated with the \emph{R} locus and that
our failure to identify the most significantly associated \emph{k}-mers as
causal variants is due to the coexistence of several variants that can cause
loss of function at the \emph{R} locus.  Still, we observed two causal variants
previously documented by Gillman et al. \cite{gillman2011} in our dataset (Additional file 1: Figures
S23 and S24).

@@

We conducted a GWAS analysis comparing accessions with normal and semi-sparse
pubes\-cence in order to detect loci associated with pubescence density. This analysis
detected signals at the \textit{Ps} locus using all three approaches (Figure
\ref{pubescence-density-main-figure}a-b and Additional file 1: Figures S25, S26
and S27). The causal
variant at this recently cloned locus is a copy number variant (CNV) overlapping the
Glyma.12g187200 gene, with higher copy number resulting in reduced pubescence \citep{liu2020ps}.
We identified a region of $\sim$42 kb that varied in copy number across
the population and was associated with the trait (Figure
\ref{pubescence-density-main-figure}c), and therefore possibly corresponded to the
25.6-kb region identified as the causal CNV by Liu et al. \cite{liu2020ps}. We found that
reads containing the most significantly associated \textit{k}-mer mapped to either of the two ends
of the CNV (Figure \ref{pubescence-density-main-figure}d). This suggests that
the most associated \textit{k}-mer is indeed linked to the CNV
that is causal at this locus, while \textit{k}-mers within the
CNV region may represent differences in sequence between the copies.

The difference in size between the CNV reported by Liu et al. \cite{liu2020ps} and our
results may be due to the presence of some unresolved sequence (a stretch of
4,635 N nucleotides) over that interval in the reference sequence that we used.
Because our SV filtering process removed any alternate allele containing N nucleotides,
we reasoned that the CNV may have been found by our pipeline but
filtered out later on. Indeed, we found a 42.8-kb duplication corresponding
to the one described above among the genotype calls made by smoove. An association
analysis of that single variant using GAPIT found a \textit{p}-value of 1.9e-23,
which would have made it the most significant SV. In this particular case, the
failure of the SV-based GWAS analysis to find identify the causal variant was due
to our filtering parameters and not to limitations of the SV genotype calls per se.

@@

We conducted a GWAS analysis comparing dull and shiny seed coat phenotypes for
the seed coat luster trait. Seed coat luster has been shown by
Gijzen et al. \cite{gijzen1999} to be largely caused by the deposition of a hydrophobic
protein (HPS) at the seed surface. Copy number variation of this gene at the
\textit{B} locus was later shown to explain variation in luster, although
sequence variation in the HPS sequence may also play a role \citep{gijzen2006}.
Using a sequencing depth-based analysis, we identified a 31-kb region spanning
positions 9,386,109-9,417,431 on chromosome Gm15 that exhibited variation in
copy number across accessions. This CNV overlapped four genes annotated as
containing HPS domains and therefore likely represented the causal variant
(Additional file 1: Figures S28 and
S29).  We found signals overlapping this
CNV using all approaches, and the most significantly associated \textit{k}-mer
was notably located within the CNV region (Additional file 1: Figures
S28 and
S30). However, we were not able to
directly link any of the variants or \textit{k}-mers identified to the causal
CNV.

@@

Both the \textit{L1} and the \textit{L2} loci for pod color have yet to be
cloned. While He et al. \cite{he2015} proposed the gene Glyma.19g101700 as likely
corresponding to \textit{L1} based on fine mapping and expression analysis, our
results as well as those reported by Bandillo et al. \cite{bandillo2017} do not support this
gene as a candidate due to its location $\sim$3 Mb away from GWAS signals. Our
analyses based on \textit{k}-mers and SNPs found signals overlapping 14 and
6 genes, respectively (Additional file 1: S31 and
S32). Of these, a MATE transporter (Glyma.19g120300)
appears as a prime candidate as the most significant \emph{k}-mer was observed
within its sequence and it is highly expressed in pods according to data hosted
on SoyBase \citep{grant2010, severin2010}. Moreover, the role of some MATE
proteins in transporting flavonoids has been demonstrated \citep{chen2015}.

@@

The two known loci for pubescence form in soybean, \textit{Pa1} and
\textit{Pa2}, have also not been cloned yet. Nevertheless, Gilbert \cite{gilbert2017}
has suggested Glyma.12g213900 as a candidate for \textit{Pa1} based on data
published by Bandillo et al. \cite{bandillo2017}. This gene encodes a MYB transcription factor
annotated as playing a role in trichome branching. Indeed, the most significant
signals found by all three approaches were located near this gene (Figure
\ref{pubescence-form-main-figure}a-b, and Additional file 1: Figures S34 and
S35).  The analyses based on SNPs and
\emph{k}-mers identified two highly significant non-synonymous SNPs in
Glyma.12g213900 that were associated with appressed pubescence (Figure
\ref{pubescence-form-main-figure}c-d and Additional file 1: Figures S36 and
S37). The alternative alleles for these two
SNPs co-occurred together in our dataset and may represent causal variants at
the \emph{Pa1} locus. The signals observed at the \emph{Pa2} locus were much
weaker and the signal observed from the \emph{k}-mer GWAS overlapped a total of
20 genes (Additional file 1: Figures S34 and
S38), none of which appeared obviously linked
to pubescence form.

@@

We used corrected dry weight (CDW) as described by de Ronne et al. \cite{deronne2022} as a
measure of horizontal resistance to \emph{P. sojae}. We found a strong signal
overlapping the genomic region reported by these authors on Gm15 using all three GWAS
approaches (Additional file 1: Figure S40). de Ronne et al. \cite{deronne2022}
proposed Glyma.15g217100 as a candidate gene for explaining phenotypic
variation at that locus based on a combination of functional annotation and
contrasting gene expression in resistant and susceptible lines four days after
inoculation. Although the signals found by our analyses did overlap
Glyma.15g217100, we found that this gene was located at the distal end of the signal
in a region with markers or \emph{k}-mers showing a weaker
association (Additional file 1: Figure S41). The
large size of the observed signal means that the gene associated with the phenotype could
be located almost anywhere in this region. However, we identified a particularly
interesting region defined by the top 1\% associations in the SNP-based
analysis. This region overlaps only three annotated genes and also contains the
most significant \emph{k}-mer (Additional file 1: Figure
S41). One of the genes in this region is
a calcium ion binding protein with an EF-hand domain (Glyma.15g217700).
Given the important role played by calcium signaling in defense
responses \citep{zhang2014}, we suggest Glyma.15g217700 as an interesting
candidate gene for explaining resistance to \emph{P. sojae} in this population.

@@

As would be expected of any GWAS analysis, the success of the \emph{k}-mer approach
depends on the quality of the phenotypic data and the complexity of the genetic
architecture of the trait. It is clear that the loci for which the
\emph{k}-mer approach was most successful (\emph{W1}, \emph{Td}, \emph{G} and
\emph{Ps}) have a simple genetic architecture and phenotypes that can be easily
determined. For other traits, challenges in determining accurate phenotypes may
result in a decreased ability to identify causal variants or genes. For
example, Bandillo et al. \cite{bandillo2017} noted that observed phenotypes for stem
termination type in the GRIN database may not correspond to the expected
genotypes at classical loci, and that the intermediate phenotype for seed coat
luster is not always consistent across different environments. For qualitative
traits, the numerical scale used to recode phenotypes for use in GWAS is somewhat
arbitrary and may influence the final results. In the case of quantitative
traits, the usual limitations to the precision and stochastic variation of
phenotypic values will necessarily make the interpretation more difficult than
for simple qualitative traits. Notwithstanding those limitations,
\textit{k}-mer-based GWAS clearly performed better than any of the other
approaches tested here given the same phenotypic data. This is partly due
to the fact that the \textit{k}-mer approach encompasses all variant types.
Moreover, it is likely that the simpler workflow that results from using
\textit{k}-mer presence/absence avoids some of the errors introduced during
variant calling and genotyping, and therefore yields more accurate results.

@@

The complexity of analyzing the results of \emph{k}-mer-based GWAS is another
limitation of this approach. Although the method developed by
Voichek and Weigel \cite{voichek2020} is in itself relatively simple and computationally
inexpensive, downstream analysis and interpretation of the output is
challenging. Indeed, there are no state-of-the-art methods for identifying the
putative genomic coordinates of a \emph{k}-mer, grouping significant
\emph{k}-mers together into coherent signals, and linking them to biologically
meaningful sequence variation. Going forward, mapping the \emph{k}-mers or
reads to a graph-based genome \citep[e.g. using the vg toolkit;][]{siren2021}
may represent an improvement to using a single linear genome.  Grouping
significant \emph{k}-mers by LD instead of by genomic position, as was done for
some of the analyses in this paper, could prove useful in determining how many
loci control a trait. In this study, we used an assembly-based approach similar
to what others have previously done \citep[e.g.][]{voichek2020, rahman2018}.
This approach allowed us to link \emph{k}-mers to their underlying variation
for simple cases involving SNPs and indels, but we were unable to link
significantly associated \emph{k}-mers to more challenging causal variants such
as those explaining the \emph{I} or \emph{B} loci. In this context, developing
approaches to systematically link significant \emph{k}-mers to the sequence
variation that underlies their presence is much needed. Furthermore, finding
ways to limit the number of spurious associations found for some traits, as was
the case for flower color and pubescence color in our study, is also needed in
order to obtain robust results.

@@

Given the current performance of SV genotyping programs, it is not clear
whether using SVs for GWAS provides any advantage over using \emph{k}-mer- or
SNP-based GWAS, or a combination of both.  Indeed, in the single case where the
SV-based analyses identified the causal variant, the \emph{k}-mer-based GWAS
identified it as well. Moreover, SNP-based GWAS provided better guidance
in delimiting the signals found at a given locus because of the much higher
density of markers compared to SVs.  The usefulness of SV-based GWAS compared
to SNP-based GWAS will largely depend on whether SVs can represent variation
that is not otherwise in LD with SNPs. In tomato, Domínguez et al. cite{dominguez2020} found
that most transposable element insertions associated with phenotypic variation
in agronomic traits were not tagged by SNPs. On the other hand, in humans,
Maretty et al. \cite{maretty2017} found that the majority of SVs were in high LD with SNPs. It
is therefore likely that how well SVs can be tagged by SNPs will depend on the
particular type of SV and on the species. A more efficient approach than
SV-based GWAS may be to conduct \emph{k}-mer-based GWAS and only then link
known SVs to significant \emph{k}-mers.  Alternatively, \textit{k}-mer-based
GWAS can be conducted specifically for \textit{k}-mers that are associated with
SVs \citep{jayakodi2020}.  Previous studies have obtained noteworthy results by
using SVs for GWAS analyses \citep[e.g.][]{zhang2015, akakpo2020,
dominguez2020}, however it would be worthwhile to see whether these results
could be replicated using \emph{k}-mer-based GWAS.

@@

We selected 389 inbred \emph{G. max} accessions based on the availability of
Illumina WGS data and phenotypic data for resistance to the oomycete \emph{P.
sojae} \citep{deronne2022}.  We identified 741 SRA runs corresponding to
Illumina paired-end sequencing data for those 389 accessions \citep[data from][]{zhou2015, valliyodan2016, fang2017, bayer2021}
and extracted paired-end reads using the fastq-dump command v. 2.9.6
\cite{sratoolkit} with the
option \texttt{-\-split-3}.
Reads that were not paired in the raw sequencing data were not used
for downstream analyses.  The runs downloaded from the SRA and associated
metadata are listed in Additional file 2.

We first filtered the sequencing data for quality and the presence of sequencing
adapters using bbduk from BBtools v. 38.25
\cite{bbtools}. Reads
were then mapped to the reference assembly version 4 of \emph{G. max}
cultivar Williams82 \citep{valliyodan2019} concatenated with chloroplast
and mitochondrion sequences obtained from SoyBase \citep{grant2010}. We used
bwa-mem v. 0.7.17-r1188 \citep{li2009-bwa} with default parameters to map the
reads in paired-end mode. Reads that were left unpaired following adapter- and
quality-trimming were aligned separately in single-end mode with bwa-mem and
merged with mapped paired-end reads using samtools \texttt{merge}
v. 1.12 \citep{li2009-samtools}.  We then added read groups identifying
individual SRA runs to the BAM files and merged the reads belonging to the same
accession using bamaddrg
\cite{bamaddrg}. We used the
resulting BAM files for all downstream analyses requiring aligned reads.
Metadata on mapping depth of aligned data can be found in Additional file 3.
